Still Missing a Cool Title

### 2006-05-30 talk/dls2006/paper.tex
Minor latexifications.

pypy-c
    The simplest variant: translated to C code with no explicit memory
    management, and linked with the Boehm conservative GC [Boehm].

pypy-c-thread
    The same, with OS thread support enabled.  (For measurement purposes,
    thread support is kept separate because it has an impact on the GC
    performance.)

pypy-c-stackless
    The same as pypy-c, plus the "stackless transformation" step which
    modifies the flow graph of all functions in a way that allows them
    to save and restore their local state, as a way to enable coroutines.

pypy-c-gcframework
    In this variant, the "gc transformation" step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

pypy-c-stackless-gcframework
    This variant combines the "gc transformation" step with the
    "stackless transformation" step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case - 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

pypy-llvm-c
    The same as pypy-c, but using the LLVM back-end instead of the C
    back-end.  The LLVM assembler-compiler gives the best results when -
    as we do here -- it optimizes its input and generates again C code,
    which is fed to GCC.

pypy-llvm-c-prof
    The same as pypy-llvm-c, but using GCC's profile-driven
    optimizations.

The speed difference with CPython 2.4.3 can be explained at two levels.
One is that CPython is hand-crafted C code that has been continuously
optimized for a decade now, whereas the Python interpreter of PyPy first
seeks flexibility and high abstraction levels.  The other, probably
dominant, factor is that various indices show that our approach places a
very high load on the GC and on the memory caches of the machine.  The
Boehm GC is known to be less efficient than more customized approach;
kernel-level profiling shows that pypy-c typically spends 30% of its
time in the Boehm library.  Our current, naively simple mark-and-sweep
GC is even quite worse.  The interaction with processor caches is also
hard to predict and account for; in general, we tend to produce
relatively large amounts of code and prebuilt data.

>>>>

{\bf pypy-c.}
    The simplest variant: translated to C code with no explicit memory
    management, and linked with the Boehm conservative GC [Boehm].

{\bf pypy-c-thread.}
    The same, with OS thread support enabled.  (For measurement purposes,
    thread support is kept separate because it has an impact on the GC
    performance.)

{\bf pypy-c-stackless.}
    The same as pypy-c, plus the "stackless transformation" step which
    modifies the flow graph of all functions in a way that allows them
    to save and restore their local state, as a way to enable coroutines.

{\bf pypy-c-gcframework.}
    In this variant, the "gc transformation" step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

{\bf pypy-c-stackless-gcframework.}
    This variant combines the "gc transformation" step with the
    "stackless transformation" step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case - 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

{\bf pypy-llvm-c.}
    The same as pypy-c, but using the LLVM back-end instead of the C
    back-end.  The LLVM assembler-compiler gives the best results when -
    as we do here -- it optimizes its input and generates again C code,
    which is fed to GCC.

{\bf pypy-llvm-c-prof.}
    The same as pypy-llvm-c, but using GCC's profile-driven
    optimizations.

The speed difference with CPython 2.4.3 can be explained at two levels.
One is that CPython is hand-crafted C code that has been continuously
optimized for a decade now, whereas the Python interpreter of PyPy first
seeks flexibility and high abstraction levels.  The other, probably
dominant, factor is that various indices show that our approach places a
very high load on the GC and on the memory caches of the machine.  The
Boehm GC is known to be less efficient than more customized approach;
kernel-level profiling shows that pypy-c typically spends 30\% of its
time in the Boehm library.  Our current, naively simple mark-and-sweep
GC is even quite worse.  The interaction with processor caches is also
hard to predict and account for; in general, we tend to produce
relatively large amounts of code and prebuilt data.
